Constraining gravitational wave amplitude birefringence with GWTC-3

### 2022-11-22 src/tex/ms.tex
New content

\begin{abstract}
    One of the major problems that physicists are facing nowadays is Einstein's theory of general relativity (GR) does not agree with quantum theories at some length scales.
    In recent decades, theorist has been working on many beyond-GR theories to unify both of them.
    Some of these theories such as Chern-Simons gravity suggest that there is gravitational wave (GW) amplitude birefringence.
    In our study, we perform parameter estimation (PE) to constrain the strength of GW amplitude birefringence.
    Compare to similar previous studies on the same topic, we perform PE on more events, including events in the third LIGO-Virgo catalog (GWTC-3), and we used a more accurate model of birefringence to describe the phenomenon.
    % results
\end{abstract}

@@

% Constraint on $\kappa$
We then calculate the constraint on $\kappa$ from the posterior of $\mu$ and $\sigma$.
% constraint

% Comparison with previous studies
In \citet{Maria_2021}, the authors gave a constraint on GW amplitude birefringence by performing statistical analysis on GWTC-2.
We convert the constraint to the same format as ours to make comparison.
They were able to constrain $\kappa$ to be $|\kappa| \lesssim 0.74$ at $1 \sigma$.

% Violin plot
In figure \ref{fig:violin_kappa}, we show the posterior of $\kappa$ for each events in the form of violin plot.
There are some events with $\kappa$ distribution deviate from $\kappa=0$.
Also, there are some events with bimodal $\kappa$ distribution.
We are going to discuss some of these events in the following.
\begin{figure*}[ht]
    \script{violin_kappa.py}
    \includegraphics[width=\textwidth]{figures/violin_kappa.pdf}
    \caption{
        The violin plot shows the posterior of $\kappa$ for all 69 events included in this study.
        Each violin represents a different event.
        The violins are sorted by the quotient of median and standard deviation of the posterior.
        Blue horizontal line represent the median value of $\mu$, and blue region represent 1 $\sigma$ confidence level.
        Orange horizontal line represent $\kappa=0$.
        This plot shows that there are some events with $\kappa$ distribution significantly deviate from $0$.
    }
    \label{fig:violin_kappa}
\end{figure*}

@@

% Case: GW150914 (example of broken degeneracy)
\subsection{Result on GW150914}

\begin{figure}[h]
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE done by us with the frequency independent birefringence and the frequency dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency independent birefringence create a degeneracy between $\kappa$ and $\iota$, while the frequency dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

Some events have bimodal $\kappa$ distribution, but before we show a bimodal event, we should understand the mechanism of a unimodal $\kappa$ distribution.
Consider GW150914, the first GW detected by LIGO.
In figure \ref{fig:corner_GW150914}, we show the posteriors of the parameters of GW150914.
With the frequency independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
To reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$, or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.
As a result, many pairs of $\iota$ and $\kappa$ can be plausible with the frequency independence.

@@

% Case: GW190521 (massive BBH)
\subsection{Result on GW190521}
\begin{figure}[h]
    \script{corner_GW190521.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW190521.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW190521.
        Colors in the plot are the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3} and the PE done by us with the frequency dependent birefringence.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        This plot shows that frequency dependent birefringence cannot break the degeneracy between $\kappa$ and $\iota$.
    }
    \label{fig:corner_GW190521}
\end{figure}

@@

>>>>

\begin{abstract}
    One of the major problems that physicists are facing nowadays is Einstein's theory of general relativity (GR) does not agree with quantum theories at some length scales.
    In recent decades, theorist has been working on many beyond-GR theories to unify both of them.
    Some of these theories such as Chern-Simons gravity suggest that there is gravitational wave (GW) amplitude birefringence.
    In our study, we perform parameter estimation (PE) to constrain the strength of GW amplitude birefringence.
    Compare to similar previous studies on the same topic, we perform PE on more events, including events in the third LIGO-Virgo catalog (GWTC-3), and we used a more accurate model of birefringence to describe the phenomenon.
    We were able to obtain a constrain an order of magnitude tighter than previous studies which used GWTC-2.
\end{abstract}

@@

% Constraint on $\kappa$
We then calculate the constraint on $\kappa$ from the posterior of $\mu$ and $\sigma$.
We generated a Gaussian distribution for each sample of $\mu$ and $\sigma$.
The Gaussian distributions were then added up to obtain the population distribution of $\kappa$.
The mean of the population distribution of $\kappa$ is \variable{output/kappa_mean.txt}.
Note that GR is recovered if $\kappa$ is $0$.

% Comparison with previous studies
In \citet{Maria_2021}, the authors gave a constraint on GW amplitude birefringence by performing statistical analysis on GWTC-2.
We convert the constraint to the same format as ours to make comparison.
They were able to constrain $\kappa$ to be $|\kappa| \lesssim 0.74$ at $1 \sigma$.
We were able to obtain a tighter constraint on $\kappa$ with $|\kappa| \lesssim 0.04$ at $1 \sigma$.
This is a order of magnitude improvement in constraining $\kappa$.
This is due to the fact that we have much more events from GWTC-3 compared to GWTC-2.

% Violin plot
In figure \ref{fig:violin_kappa}, we show the posterior of $\kappa$ for each events in the form of violin plot.
There are some events with $\kappa$ distribution deviate from $\kappa=0$.
Also, there are some events with bimodal $\kappa$ distribution.
We are going to discuss some of these events in the following.

\begin{figure*}[ht]
    \script{violin_kappa.py}
    \includegraphics[width=\textwidth]{figures/violin_kappa.pdf}
    \caption{
        The violin plot shows the posterior of $\kappa$ for all 69 events included in this study.
        Each violin represents a different event.
        The violins are sorted by the quotient of median and standard deviation of the posterior.
        Blue horizontal line represent the median value of $\mu$, and blue region represent 1 $\sigma$ confidence level.
        Orange horizontal line represent $\kappa=0$.
        This plot shows that there are some events with $\kappa$ distribution significantly deviate from $0$.
    }
    \label{fig:violin_kappa}
\end{figure*}

@@

% Case: GW150914 (example of broken degeneracy)
\subsection{Result on GW150914}
Some events have bimodal $\kappa$ distribution, but before we show a bimodal event, we should understand the mechanism of a unimodal $\kappa$ distribution.
Consider GW150914, the first GW detected by LIGO.
In figure \ref{fig:corner_GW150914}, we show the posteriors of the parameters of GW150914.
With the frequency independent birefringence model, the posteriors for $\cos\iota$ look different from the posteriors assuming GR.
This is because, for a nonprecessing system, there is a degeneracy between $\kappa$ and $\iota$ if the frequency dependence is not included.
To reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$, or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.
As a result, many pairs of $\iota$ and $\kappa$ can be plausible with the frequency independence.

@@

\begin{figure}[h]
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE done by us with both frequency independent and dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency independent birefringence create a degeneracy between $\kappa$ and $\iota$, while the frequency dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

% Case: GW190521 (massive BBH)
\subsection{Result on GW190521}

@@

\begin{figure}[h]
    \script{corner_GW190521.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW190521.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW190521.
        Colors in the plot are the PE with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3} and the PE done by us with the frequency dependent birefringence.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        This plot shows that frequency dependent birefringence cannot break the degeneracy between $\kappa$ and $\iota$.
    }
    \label{fig:corner_GW190521}
\end{figure}
